\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-01-06 paper.tex

\includegraphics[width=1\linewidth,height=0.3\textheight]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.3\textheight]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of UMAP layout
of the two C-shaped clusters data, using three different bin inputs: (a)
\(b = 80 \text{ } (10, \text{ }8)\), (b)
\(b = 130 \text{ } (13, \text{ }10)\), and (c)
\(b = 391 \text{ } (23, \text{ }17)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. At the
smallest bin size, the data structure is discontinuous, suggesting that
there are too many bins. Using the MSE of the model fit in
\(7\text{-}D\) helps decide on a useful choice of number of bins.}

@@

\includegraphics[width=1\linewidth,height=0.6\textheight]{paper_files/figure-pdf/fig-two_non_linear_diff_shaped_close_clusters-mse-1.pdf}

@@

{\centering \includegraphics[width=1\linewidth,height=0.9\textheight]{paper_files/figure-pdf/best-fit-umap-tsne-pacamp-1.pdf}

@@

\caption{\label{fig-five-gau-projs}Model in \gD{}, on the layouts a1,
b1, and c1, and two views of the fit in projections of each layout from
\(4\text{-}D\), for the five Gaussian cluster data. All \gD{} models
show five well-separated clusters with different gaps. Layout a1 shows
small separations, while layout b2 shows large gaps. On the other hand,
layout b3 shows average gaps with exact positions as in \(4\text{-}D\).
Assessing the different models fit for different layouts for a specific
cluster (highlighted in red) shows filled-out shape in a1, a2, and b1,
b2. On the other hand, c1 and c2 shows pancake shape. Videos of the
langevitour animations are available at
\url{https://youtu.be/5Y1hE4i7N2k}, \url{https://youtu.be/5Y1hE4i7N2k},
and \url{https://youtu.be/5Y1hE4i7N2k} respectively.}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-1.pdf}

@@

\subsubsection{scDEED}\label{scdeed}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

@@

\includegraphics[width=1\linewidth,height=1\textheight]{paper_files/figure-pdf/fig-mnist-mse-1.pdf}

@@

\includegraphics[width=1\linewidth,height=0.8\textheight]{paper_files/figure-pdf/fig-model-error-mnist-1.pdf}

>>>>

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-hex-param-1.pdf}

@@

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-p-d-error-in-2d-two-curvy-1.pdf}

@@

\caption{\label{fig-bins-two-curvy}Hexbin density plots of tSNE layout
of the two C-shaped clusters data, using three different bin inputs: (a)
\(b = 80 \text{ } (10, \text{ }8)\), (b)
\(b = 130 \text{ } (13, \text{ }10)\), and (c)
\(b = 391 \text{ } (23, \text{ }17)\). Color indicates standardized
counts, dark indicating high count and light indicates low count. At the
smallest bin size, the data structure is discontinuous, suggesting that
there are too many bins. Using the MSE of the model fit in
\(7\text{-}D\) helps decide on a useful choice of number of bins. XXX
All look the same}

@@

\includegraphics[width=0.6\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-two_non_linear_diff_shaped_close_clusters-mse-1.pdf}

@@

{\centering \includegraphics[width=0.9\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/best-fit-umap-tsne-pacamp-1.pdf}

@@

\caption{\label{fig-five-gau-projs}NLDR's organise points in the \gD{}
layout in different ways, possibly misleadingly, illustrated using three
layouts: (a) t-SNE, (b) UMAP, (c) PaCMap. The data has five Gaussian
cluster in \(4\text{-}D\). The bottom row of plots shows a \gD{}
projection from a tour on \(4\text{-}D\) revealing the differences
generated by the layouts on the model fits. We would expect the model
fit to be like that in (a2) where it is distinctly is separate for each
cluster but like a hairball in each. This would indicate the distinct
clusters, each being fully \(4\text{-}D\). With (c2), the curiousity is
that the model is a \gD{} pancake shape in \(4\text{-}D\), indicating
that there is some ordering of points done by PaCMAP, posisbly along
some principal component axes. Videos of the langevitour animations are
available at \url{https://youtu.be/oQxEb4wRdHI},
\url{https://youtu.be/JW49csPpDx4}, and XXX respectively.}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-pbmc-mse-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-model-pbmc-author-proj-1.pdf}

@@

\subsubsection{Comparison with results of scDEED
recommendations}\label{comparison-with-results-of-scdeed-recommendations}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-pbmc-mse-umap-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-model-pbmc-author-proj-scdeed-1.pdf}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-mnist-mse-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-model-error-mnist-1.pdf}
